---
title: "custom_recipes"
format: pdf
editor: visual
---

# Custom Recipes

We will load a phylogenetic tree in Newick format for demonstration:

Prepare session:

```{r, message=FALSE, warning=FALSE}
library(ggtree)
library(treeio)
library(ggplot2)
library(dplyr)
```

```{r}
# Load a tree file
tree <- read.newick("example_2_phylo.nwk")
tree$node.label <- round(as.numeric(tree$node.label), digits = 3)
```

## Circular Layout Visualization

Circular layouts are excellent for presenting large or complex trees. They allow for efficient space usage and a visually appealing presentation.

```{r}
# Circular tree with basic styling
circular_tree <- ggtree(tree, layout = "circular", 
                        branch.length = "none") +
  geom_tiplab(size = 2, aes(angle = angle), offset = 0.5) +
  geom_nodelab(color = "black", size = 2, nudge_y = 0.5) 

circular_tree
```

### Highlighting Clades in Circular Trees

Highlighting specific clades can make the tree more interpretable. Use the `geom_hilight` function:

```{r}
# Highlight specific clades
highlight_tree <- circular_tree +
  geom_hilight(node = 199, fill = "blue", alpha = 0.5) +
  geom_hilight(node = 200, fill = "lightgreen", alpha = 0.5)

highlight_tree
```

### Adding Heatmaps to Circular Trees

We can add metadata as heatmaps around the tree:

```{r, message=FALSE}
# Cargar metadata
metadata <- readr::read_tsv("Experiment_Design_example_2.tsv")
metadata <- metadata %>%
  rename(label = Assembly.Accession)  

# The rows have to be in the same order as the tree
metadata_ordered <- metadata %>%
  filter(label %in% tree$tip.label) %>%
  arrange(match(label, tree$tip.label))

metadata_ordered$Phylogroup <- replace(metadata_ordered$Phylogroup, is.na(metadata_ordered$Phylogroup), "NA")

metadata_ordered <- metadata_ordered[,c(-1,-2)]
rownames(metadata_ordered) <- make.unique(metadata_ordered$Strain)
colnames(metadata_ordered)[1] <- "label"

# Change tip label
tree$tip.label <- metadata_ordered$label

# Merge metadata with tree
p_circular_heatmap <- ggtree(tree, layout = "circular", 
                        branch.length = "none") %<+% metadata_ordered +
  geom_tiplab(size = 2, aes(angle = angle), offset = 0.5) +
  geom_nodelab(color = "black", size = 2, nudge_y = 0.5)+
  geom_tippoint(aes(color = source), size = 2) +
  scale_color_manual(values = c("darkred", "darkcyan")) +
  geom_hilight(node = 199, fill = "blue", alpha = 0.5) +
  geom_hilight(node = 200, fill = "lightgreen", alpha = 0.5)

p_circular_heatmap
```

## Styling Trees with Advanced Themes

`ggtree` provides options to customize the appearance of trees using themes. Here's how to use them effectively.

### Customizing Themes for Trees

```{r}
# Applying a theme with a clean background
styled_tree <- p_circular_heatmap +
  theme_tree2() +
  theme(
    plot.background = element_rect(fill = "darkgrey", color = NA),
    panel.border = element_blank(),
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )

styled_tree
```

## Find node-tip association

If you want to fill color some nodes using the tip label, use this:

```{r}
# Extract the tree data
tree_data <- as_tibble(tree)

# Select the labels to search
# In this case use the labels that have data in Patotype
strains <- metadata_ordered %>% 
  filter(!is.na(Patotype)) %>% 
  select(label) %>% 
  as.vector() %>% unlist()

# Find tips and their parent nodes
tips_and_nodes <- tree_data %>% 
  select(label, parent) %>% 
  filter(label %in% strains)


p_circular_heatmap +
  geom_hilight(node = tips_and_nodes$parent, fill = "green", alpha = 0.5) 
```

### Adding Multiple Aesthetic Layers

Combine multiple layers for richer visuals:

```{r}
# Adding branches with customized colors
pal <- c(
  RColorBrewer::brewer.pal(6, "Set1"),
  "#FFFFFF"
)

p <- ggtree(tree, ladderize = TRUE, branch.length = "none", layout = "circular") %<+% metadata_ordered + 
  geom_tiplab(aes(label = label, angle = angle), linesize=.1, size = 4, align = FALSE, color = "black",
              offset = 1.5) +
  geom_nodelab(size = 4, color = "black", nudge_x = 0.1) +
  geom_tippoint(aes(color = Phylogroup), size = 5, shape = 20) + 
  geom_tippoint(aes(shape = source), size = 4, alpha = 0.5) + 
  scale_shape_manual(values = c(1, 8)) + labs(shape = "Source") +
  scale_color_manual(values = pal) +
  geom_hilight(node = tips_and_nodes$parent, fill = "green", alpha = 0.5) 

p
```

## Add heatmap

```{r, message=FALSE}
matrix_data <- data.frame(Patotype = metadata_ordered$Patotype)
rownames(matrix_data) <- make.unique(metadata_ordered$label)

# Add heatmap
gheatmap(p, matrix_data, width=0.05, 
         colnames=FALSE, legend_title="Patotype", color = NA, ) +
  theme(text = element_text(size = 20))
```

## Combining Trees with ggnewscale

Use the `ggnewscale` package to add multiple scales of colors:

```{r}
# Adding new scale for different metadata layers
multi_scale_tree <- ggtree(tree, layout = "circular", branch.length = "none") %<+% metadata_ordered +
  geom_tippoint(aes(color = Patotype), size = 5) +
  scale_color_brewer(palette = "Set2", name = "Patotype") +
  ggnewscale::new_scale_color() +
  geom_tippoint(aes(fill = Phylogroup), shape = 21, size = 3, color = "black") +
  scale_fill_brewer(palette = "Set1", name = "Phylogroup") +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    plot.title = element_text(size = 14, hjust = 0.5, face = "bold")
  ) +
  ggtitle("Tree with Multiple Aesthetic Layers")

multi_scale_tree
```
