---
title: "ggtree_custom_recipes"
format: pdf
editor: visual
---

# `ggtree` Custom Recipes

We will load a phylogenetic tree in Newick format for demonstration.

Prepare session:

```{r, message=FALSE, warning=FALSE}
library(ggtree)
library(treeio)
library(ggplot2)
library(dplyr)
```

## 

```{r}
# Load a tree file
tree <- read.newick("data/example_2_phylo.nwk")
tree$node.label <- round(as.numeric(tree$node.label), digits = 3)
```

## Circular Layout Visualization

Circular layouts are excellent for presenting large or complex trees. They allow for efficient space usage and a visually appealing presentation.

```{r}
# Circular tree with basic styling
circular_tree <- ggtree(tree, layout = "circular", 
                        branch.length = "none") +
  geom_tiplab(size = 2, aes(angle = angle), offset = 0.5) +
  geom_nodelab(color = "black", size = 2, nudge_y = 0.5) 

circular_tree
```

### Highlighting Clades in Circular Trees

Highlighting specific clades can make the tree more interpretable. Use the `geom_hilight` function:

```{r}
# Highlight specific clades
highlight_tree <- circular_tree +
  geom_hilight(node = 199, fill = "lightblue", alpha = 0.5) +
  geom_hilight(node = 200, fill = "lightgreen", alpha = 0.5)

highlight_tree
```

### Adding Heatmaps to Circular Trees

We can add metadata as heatmaps around the tree:

```{r}
# Cargar metadata
metadata <- readr::read_tsv("data/Experiment_Design_example_2.tsv")

colnames(metadata)[1] <- "label"

# The rows have to be in the same order as the tree
metadata_ordered <- metadata %>%
  filter(label %in% tree$tip.label) %>%
  arrange(match(label, tree$tip.label))

metadata_ordered$Phylogroup <- replace(metadata_ordered$Phylogroup, is.na(metadata_ordered$Phylogroup), "NA")



#Convert tree information to tibble object and add metadata
treetib <- as_tibble(tree)
treetib <- left_join(treetib, metadata_ordered, by = c('label'))

tree_ec <- as.treedata(treetib)




# Merge metadata with tree
p_circular_heatmap <-  ggtree(tree_ec, layout = "circular", 
                        branch.length = "none") +
  geom_tiplab(size = 2, aes(label = Strain,angle = angle), offset = 0.5) +
  geom_nodelab(color = "black", size = 2, nudge_y = 0.5)+
  geom_tippoint(aes(color = source), size = 2) +
  scale_color_manual(values = c("darkred", "darkcyan"))

p_circular_heatmap
```

## Styling Trees with Advanced Themes

`ggtree` provides options to customize the appearance of trees using themes. Here's how to use them effectively.

### Customizing Themes for Trees

```{r}
# Applying a theme with a clean background
styled_tree <- circular_tree +
  theme_tree2() +
  theme(
    plot.background = element_rect(fill = "#f9f9f9", color = NA),
    panel.border = element_blank(),
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )

styled_tree
```

### Adding Multiple Aesthetic Layers

Combine multiple layers for richer visuals:

```{r}
# Adding node points and branches with customized colors
layered_tree <- ggtree(tree_ec, layout = "circular", branch.length = "none") +
  geom_tree(color = "darkgrey", size = 0.6) +
  geom_tippoint(aes(color = Phylogroup), size = 3) +
  scale_color_viridis_d() +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    plot.title = element_text(size = 14, hjust = 0.5, face = "bold")
  ) +
  ggtitle("Layered Circular Tree Visualization")

layered_tree
```

## Combining Trees with ggnewscale

Use the `ggnewscale` package to add multiple scales of colors:

```{r, warning=FALSE}
# Adding new scale for different metadata layers
multi_scale_tree <- ggtree(tree_ec, layout = "circular", branch.length = "none") +
  geom_tippoint(aes(color = Patotype), size = 2) +
  scale_color_brewer(palette = "Set2", name = "Category") +
  ggnewscale::new_scale_color() +
  geom_tippoint(aes(fill = Phylogroup), shape = 21, size = 3, color = "black") +
  scale_fill_brewer(palette = "Set1", name = "Fill Category") +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    plot.title = element_text(size = 14, hjust = 0.5, face = "bold")
  ) +
  ggtitle("Tree with Multiple Aesthetic Layers")

multi_scale_tree
```
