# Newick Format {#newick}

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

Load the neccesary libraries for this section
We will use ape to charge our phylogenetic tree and ggtree to visualize it.
```{r eval=TRUE, warning=FALSE}
library("ape") #We use this library to read the newick texts
library("ggtree")
```

Newick format is a way of representing a phylogeny (a tree) with parentheses and commas. Its basic elements are:

* Branches
* Nodes
   + Leaves: Nodes with no descendants, also called tips
   + Internal nodes: Nodes with descendants

A Minimal newick tree has only branches. There is a branch for each comma and each internal parenthesis pair.

```{r eval=TRUE, warning= FALSE}
#   (  , (,)  )
#        ^ ^    internal parenthesis pair.
#   ^         ^  external parenthesis
```

As we will see later the standard phylogeny object in R is an S3 list of class "phylo", popularized by the ape package. Lets obtain our tree in phylo format.
```{r eval=TRUE, warning= FALSE}
tree<-"(,(,));" 
tree<-read.tree(text=tree)
```

And now we can use ggtree to generate an image
```{r newick1, fig.cap='A minimal tree', out.width='50%', fig.align='center', warning=FALSE}
ggtree(tree)
```

Here you can see leaves (tips) and internal nodes colored in red and blue, respectively

```{r newick2, fig.cap='Leaves (tips) in red and internal nodes in blue', out.width='50%', fig.align='center', warning=FALSE}
ggtree(tree)+
  geom_tippoint(size=3, color="red")+
  geom_nodepoint(size=3, color="blue")
```

The tree leaves can be labeled with clade names( ej. species, strains)

```{r newick3, fig.cap='If a Newick has leave names', out.width='50%', fig.align='center', warning=FALSE}
tree<-"(cat,(dog,fox));" 
tree<-read.tree(text=tree)
ggtree(tree)+geom_tiplab()
```

You can indicate different lengths for each branch
The length reflects the amount of evolutionary change of the branch

```{r newick4, fig.cap='If a newick has branch lengths', out.width='50%', fig.align='center', warning=FALSE}
tree<-"(cat:0.6,(dog:0.1,fox:0.1):0.2);" 
tree<-read.tree(text=tree)
ggtree(tree)+geom_tiplab()
```

Internal nodes are common ancestors for a group of leaves and you can also label them.

```{r newick5, fig.cap='Newick with internal node labels', out.width='50%', fig.align='center', warning=FALSE}
tree<-"(cat:0.6,(dog:0.1,fox:0.1)Anc2:0.2)Anc1;" 
tree<-read.tree(text=tree)
ggtree(tree)+geom_tiplab()+geom_nodelab()
```

... or instead of a name you can use a bootstrap value:
Bootstrap values show how frequently we observe same branch is we repeat the generation of a phylogeny hundreds of times with random samples of our data. They are percentages.

```{r newick6, fig.cap='Newick with bootstraps', out.width='50%', fig.align='center', warning=FALSE}
#Notice the format bootstrap_value:distance in internal nodes
tree<-"(cat:0.6,(dog:0.1,fox:0.1)90:0.2)98;" 
tree<-read.tree(text=tree)
ggtree(tree,)+geom_tiplab()+geom_nodelab()
```

Here we have a more complex example
```{r newick7, out.width='50%', fig.align='center', warning=FALSE}
tree<-"(Bovine:0.69395,
      (Gibbon:0.36079,
          (Orang:0.33636,
              (Gorilla:0.17147,
                  (Chimp:0.19268, Human:0.11927)50:0.08386
               )90:0.06124
           )70:0.15057
       )80:0.54939,Mouse:1.21460)90;"
tree<-read.tree(text=tree)
ggtree(tree)+geom_tiplab()+geom_nodelab()
```





