# Newick Format {#newick}

```{r, echo=FALSE}
knitr::include_graphics(normalizePath("docs/_main_files/figure-html/"))
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

Load the neccesary libraries for this section

```{r eval=TRUE, warning=FALSE}
library("ape") #We use this library to read the newick texts
library("ggtree")
```

Newick format is a way of representing a phylogeny (a tree) with parentheses and commas. Its basic elements are:

* Branches
* Nodes
   + Leaves: Nodes with no descendants, also colled tip points
   + Internal nodes: Nodes with descendants

A Minimal newick tree has only branches. There is a branch for each comma and each internal parenthesis pair.

```{r eval=TRUE, warning= FALSE}
#   (  , (,)  )
#        ^ ^    internal parenthesis pair.
#   ^         ^  external parenthesis
```

Lets obtain this tree in phylo format
```{r eval=TRUE, warning= FALSE}
tree<-"(,(,));" 
tree<-read.tree(text=tree)
```

And now we can use ggtree to generate an image
```{r fig1, fig.cap='A minimal newick tree', out.width='50%', fig.align='center', warning=FALSE}
ggtree(tree)+geom_tiplab()
```

Here you can see tip points and internal nodes colored in red and blue, respectively

```{r eval=TRUE, warning=FALSE}
ggtree(tree)+
  geom_tippoint(size=3, color="red")+
  geom_nodepoint(size=3, color="blue")
```

The tree leaves (tip points) can be labeled 
Labels correspond to clades (species, strains)

```{r eval=TRUE, warning=FALSE}
tree<-"(cat,(dog,fox));" 
tree<-read.tree(text=tree)
ggtree(tree)+geom_tiplab()
```

You can indicate different lengths for each branch
The length reflects the amount of evolutionary change of the branch

```{r eval=TRUE, warning=FALSE}
tree<-"(cat:0.6,(dog:0.1,fox:0.1):0.2);" 
tree<-read.tree(text=tree)
ggtree(tree)+geom_tiplab()
```

You can also label internal nodes.
Internal nodes are common ancestors for a group of leaves
```{r eval=TRUE, warning=FALSE}
tree<-"(cat:0.6,(dog:0.1,fox:0.1)Anc2:0.2)Anc1;" 
tree<-read.tree(text=tree)
ggtree(tree)+geom_tiplab()+geom_nodelab()
```

... or instead of a node label you can use a bootstrap value
Bootstrap values indicate the support of a given node.
Here, the event with fox and dog having a common ancestor has a confidence of 0.9 

```{r eval=TRUE, warning=FALSE}
tree<-"(cat:0.6,(dog:0.1,fox:0.1)0.9:0.2)0.98;" 
tree<-read.tree(text=tree)
ggtree(tree,)+geom_tiplab()+geom_nodelab()
```

Here we have a more complex example

```{r eval=TRUE, warning=FALSE}
tree<-"(Bovine:0.69395,
      (Gibbon:0.36079,
          (Orang:0.33636,
              (Gorilla:0.17147,
                  (Chimp:0.19268, Human:0.11927)0.5:0.08386
               )0.9:0.06124
           )0.7:0.15057
       )0.8:0.54939,Mouse:1.21460)0.9;"
tree<-read.tree(text=tree)
ggtree(tree)+geom_tiplab()+geom_nodelab()
```

## APPENDIX1 1: Converting to a R phylo object (see apendix below)

All of the following functions can create a phylo object from a newick file
```{r eval=FALSE, warning=FALSE}
ape::read.tree("example_phylo.nwk") %>% class
phyloseq::read_tree("example_phylo.nwk") %>% class
phytools::read.newick("example_phylo.nwk") %>% class
treeio::read.newick("example_phylo.nwk") %>% class
```



