# Newick and phylo trees {#newickphylo}

## Newick files

Newick format is a way of representing a phylogeny tree with parentheses and commas. If we have a simple phylogeny like the following:

```{r eval=TRUE, warning= FALSE}   
#             ______A        A, B and C are leaves or Tip nodes.
#      _____.|               The dots mark internal nodes.
#     |      |_______B       
#    .|
#     |____________C
```

Its most simple Newick format representation would be:
```{r eval=TRUE, warning= FALSE}   
#  "(,(,));"  With no leaf labels
# 
```

As we will see later the standard phylogeny object in R is an object of class "phylo". Lets obtain our tree in phylo format.

```{r eval=TRUE, warning= FALSE}
library(ape)
tree<-"(,(,));" 
tree<-ape::read.tree(text=tree)
```

And now we can use ggtree to generate an image
```{r newick1, out.width='50%', fig.align='center', message=FALSE, warning=FALSE}
library(ggtree)
ggtree(tree)
```

Here you can see leaves (Tip nodes) and internal nodes colored in red and blue, respectively

```{r newick2, fig.cap='Leaves (tips) in red and internal nodes in blue', out.width='50%', fig.align='center', warning=FALSE}
ggtree(tree)+
  geom_tippoint(size=3, color="red")+
  geom_nodepoint(size=3, color="blue")
```

The tree leaves can be labeled with clade names( ej. species, strains)

```{r newick3, fig.cap='If a Newick has leave names', out.width='50%', fig.align='center', warning=FALSE}
tree<-"(cat,(dog,fox));" 
tree<-read.tree(text=tree)
ggtree(tree)+geom_tiplab()
```

You can indicate different lengths for each branch
The length reflects the amount of evolutionary change of the branch

```{r newick4, fig.cap='If a newick has branch lengths', out.width='50%', fig.align='center', warning=FALSE}
tree<-"(cat:0.6,(dog:0.1,fox:0.1):0.2);" 
tree<-read.tree(text=tree)
ggtree(tree)+geom_tiplab()
```

Internal nodes are common ancestors for a group of leaves and you can also label them.

```{r newick5, fig.cap='Newick with internal node labels', out.width='50%', fig.align='center', warning=FALSE}
tree<-"(cat:0.6,(dog:0.1,fox:0.1)Anc2:0.2)Anc1;" 
tree<-read.tree(text=tree)
ggtree(tree)+geom_tiplab()+geom_nodelab()
```

... or instead of a name you can use a bootstrap value:
Bootstrap values show how frequently we observe same branch is we repeat the generation of a phylogeny hundreds of times with random samples of our data. They are percentages.

```{r newick6, fig.cap='Newick with bootstraps', out.width='50%', fig.align='center', warning=FALSE}
#Notice the format bootstrap_value:distance in internal nodes
tree<-"(cat:0.6,(dog:0.1,fox:0.1)90:0.2)98;" 
tree<-read.tree(text=tree)
ggtree(tree,)+geom_tiplab()+geom_nodelab()
```

Here we have a more complex example
```{r newick7, out.width='50%', fig.align='center', warning=FALSE}
tree<-"(Bovine:0.69395,
      (Gibbon:0.36079,
          (Orang:0.33636,
              (Gorilla:0.17147,
                  (Chimp:0.19268, Human:0.11927)50:0.08386
               )90:0.06124
           )70:0.15057
       )80:0.54939,Mouse:1.21460)90;"
tree<-read.tree(text=tree)
ggtree(tree)+geom_tiplab()+geom_nodelab()
```

## phylo objects 

The standard phylogeny object in R is the *phylo* object popularized by the *ape package*. The phylo class is used by most R packages that deal with phylogenies. Thanks to that there are many functions to *create a phylo* object *from a Newick* format file:

```{r eval=FALSE, warning=FALSE}
ape::read.tree("example_eco_model.nwk") %>% class
phyloseq::read_tree("example_eco_model.nwk") %>% class
phytools::read.newick("example_eco_model.nwk") %>% class
treeio::read.newick("example_eco_model.nwk"") %>% class
```

A phylo object describes a phylogeny with the following data:
```{r, warning=FALSE, message=FALSE}
library(ape)
tree0<-ape::read.tree("data/example_eco_model.nwk")
class(tree0)
names(tree0)
```

*tip.label* shows the names or ids of the species, genomes or strains whose evolutionary history we are studying (names of the leaves or tips):
```{r, warning=FALSE, message=FALSE}
head(tree0$tip.label)
```

*node.label* has the labels of the *internal nodes*. In many cases, these tags are the bootstrap support values.

> [!NOTE]
> When we: 1) Create phylogeny from a sequence alignment with RaxML and b) calculate bootstraps for the phylogeny we obtain a final tree with bootstrap support values added as *internal node* labels.

```{r, warning=FALSE, message=FALSE}
head(tree0$node.label)
```

Phylo contains an *edges matrix* of ancestor --- > descendant connections. Here the numbers are IDs of the internal and tip nodes.

* Column 1: Ancestral nodes (closer to the root)
* Column 2: Descendant nodes.
* The leaves are numbered from 1 to the total number of leaves.
* The internal node ids go after the leave ids.

```{r, warning=FALSE, message=FALSE}
head(tree0$edge)
```

You can obtain the *label for a given node ID* like this
```{r, warning=FALSE, message=FALSE}
match("GCF_000010385.1", tree0$tip.label)
```


*edge.length* holds the lengths of the edges of the edge matrix.

> [!NOTE]
> When we create phylogeny from a sequence alignment with RaxML, branch lengths are genetic distances, defined as the expected number of nucleotide or amino acid substitutions per site.

```{r, warning=FALSE, message=FALSE}
head(tree0$edge.length)
```



