# Main ggtree functions {#ggtree1}

The ggtree package in R is a powerful tool for visualizing and annotating phylogenetic trees. It is built on top of the ggplot2 package and allows for highly customizable and publication-ready tree visualizations.

## Prerequisites

To follow this tutorial, ensure you have the following R packages installed:

```{r, eval=FALSE, message=FALSE, warning=FALSE}
install.packages("tidyverse")
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

BiocManager::install("ggtree")
BiocManager::install("treeio")
```

Load the neccesary packages

```{r, message=FALSE, warning=FALSE}
library(ggtree)
library(treeio)
library(ggplot2)
library(dplyr)
```

Loading a Phylogenetic Tree

We will use a .nwk tree for demonstration.

ggtree supports various formats of phylogenetic trees. The most common is the Newick format.


```{r, warning=FALSE, message=FALSE}
# Load tree from a Newick file as a phylo object (read.nwwick from treeio)
tree <- read.newick("data/example_2_phylo.nwk")

# Print tree structure
print(tree)
```
 
Basic Tree Visualization

Visualizing the tree is straightforward

```{r, warning=FALSE,message=FALSE}
# Basic tree plot
p <- ggtree(tree)
print(p)
```

Adding Tip Labels

You can add labels to the tips of the tree using geom_tiplab:
```{r, warning=FALSE, message=FALSE}
p <- ggtree(tree) +
  geom_tiplab()
print(p)
```

Parameters for geom_tiplab:

* size: Adjusts the size of the labels.
* align: Aligns labels horizontally.
* angle: Rotates labels (useful for circular trees).

```{r, warning=FALSE, message=FALSE}
p <- ggtree(tree) +
  geom_tiplab(size = 3, align = TRUE, angle = 45)
print(p)
```

Tree Layouts

ggtree supports various layouts for trees, such as:

* rectangular (default)
* circular
* fan
* slanted
* radial

Circular layout
```{r, warning=FALSE, message=FALSE}
p_circular <- ggtree(tree, layout = "circular")
print(p_circular)
```

Fan layout
```{r, warning=FALSE, message=FALSE}
p_fan <- ggtree(tree, layout = "fan")
print(p_fan)
```

## Annotating Trees

Adding Node Labels

To annotate nodes, use geom_nodelab:

```{r, warning=FALSE, message=FALSE}
p <- ggtree(tree) +
  geom_nodelab(aes(label = node), size = 3)
print(p)
```

Highlighting Clades

Highlight specific clades with geom_hilight:

```{r, warning=FALSE, message=FALSE}
p <- ggtree(tree) +
  geom_hilight(node = 199, fill = "blue", alpha = 0.3)
print(p)
```

Parameters for geom_hilight

* node: Specify the node ID to highlight.
* fill: The fill color for the highlight.
* alpha: Transparency of the highlight.

Annotating Specific Tips

```{r, eval=FALSE, warning=FALSE, message=FALSE}
p <- ggtree(tree) +
  geom_tippoint(aes(color = label), size = 3)+
  theme(legend.position = "none")
print(p)
```

## Adding Metadata

Metadata can be merged into the tree object and displayed, metadata is a data.frame:

```{r, message=FALSE, warning=FALSE, message=FALSE}
# Charge metadata and keep relevant columns
metadata <- readr::read_tsv("data/Experiment_Design_example_2.tsv")
metadata<-metadata %>% select(Assembly.Accession,Strain,Phylogroup,Patotype,Geographic.location,source)
print(head(metadata))
print(tail(metadata))
```
metadata <- metadata %>% rename(label = Assembly.Accession)  


```{r, message=FALSE, warning=FALSE}
# Metadata rows have to be in the same order as the tree

tree<-read.tree("data/example_2_phylo.nwk")

metadata <- metadata %>%
  filter(Assembly.Accession %in% tree$tip.label) %>%
  arrange(match(Assembly.Accession, tree$tip.label))
```

You can rename the taxa in the tree object
```{r, message=FALSE, warning=FALSE}
# Rename the taxa in the tree object
tree2 <- rename_taxa(tree, metadata, Assembly.Accession, Strain)
p<-ggtree(tree2)+geom_tiplab()
print(p)
```

Or even better, attach metadata to ggtree object ...
```{r, message=FALSE, warning=FALSE}
p <- ggtree(tree) %<+% metadata
```

... and use any metadata column as labels
```{r, message=FALSE, warning=FALSE}
p<-p + geom_tiplab(aes(label = Strain))
print(p)
```

You could also change tip labels by
```{r, message=FALSE, warning=FALSE}
tree2<-tree
tree2$tip.label <- metadata$Strain
```

Color elements using metadataa
```{r, warning=FALSE, message=FALSE}
# Plot tree with metadata
p <- ggtree(tree) %<+% metadata +
  geom_tiplab(aes(label=Strain))+
  geom_tippoint(aes(color = source), size = 3)
print(p)
```

Note: There is a metadata fusion alternative with full_join, check in the code treeio.R

## Creating Heatmaps

#Use gheatmap to add heatmaps to the tree:
#THIS CODE IS CURRENTLY FAILING!!
```{r, warning=FALSE, message=FALSE,eval=FALSE}
# Example matrix
set.seed(123)
matrix_data <- data.frame(Patotype = metadata$Patotype,
                          Geography=metadata$Geographic.location)
rownames(matrix_data) <-metadata$Assembly.Accession
# Add heatmap
gheatmap(p, matrix_data, width=0.05, colnames=FALSE, legend_title="Patotype" ) +
  theme(text = element_text(size = 20))
```


Session Info

```{r, warning=FALSE, message=FALSE,eval=FALSE}
sessionInfo()
```
