# Biological annotations {#metadata}

A phylo object is excellent to store a phylogeny structure but it was not designed to store the characteristics of the studied organisms. As we need this data to make sense of the phylogeny we store it in tables elsewhere. 

Here is a list of some frequently used annotations with emphasis in bacteria:

* Taxonomy: Species or clade within a species (ej. B2 in E. coli)
* Isolate/strain name
* Sequence type: an specific combination of marker gene alleles. (e. g. In *E. coli*: adk, fumC, gyrB, purA, recA, icd and mdh)
* Geography: Country, state
* Host type: Human, pig, cow, tomato, others.
* Group within a study: this study, outgroup, model strain, others.

To work with the corresponding annotations wee need to charge the metadata ane make its ids follow the same order as tip.labels of the phylo tree

A) Charge
```{r, message=FALSE, warning=FALSE, message=FALSE}
metadata <- readr::read_tsv("data/example_eco_model.tsv")
print(head(metadata))
```
B) Reorder

```{r, message=FALSE, warning=FALSE}
library(dplyr)

tree_eco<-read.tree("data/example_eco_model.nwk")

metadata <- metadata %>%
  filter(Assembly.Accession %in% tree_eco$tip.label) %>%
  arrange(match(Assembly.Accession, tree_eco$tip.label))

```


We will now show two easy ways to work with these metadata now, but the corresponding tools will be covered in more detail in later chapters.

### Add metadata to ggtree objects

ggtree objects are a kind of ggplot object designed to represent phylogenies. Here we construct a ggtree object from a phylo object and add the metadata to it. 

After we add the metadata we can use it to customize the plot aesthetics.
```{r, message=FALSE, warning=FALSE}

library(ggtree)

p<-ggtree(tree_eco)

p<-p %<+% metadata

p<-p + geom_tiplab(aes( label=Strain)) +
  geom_tippoint(aes(color = Phylogroup), size = 3)

p

```

### phylo to tidytree

tidytree objects are phylogenetic tree data structures transformed into a "tidy" format. They are S4 objects with a rigid encapsulation and building. Fortunately, tibble functions make the coversion from a phylo to a tidytree object straightforward. 

```{r, warning=FALSE, message=FALSE}

library(tibble)

treetib<-as_tibble(tree_eco)

class(treetib)
head(treetib)
```

We can now join the two table like objects

```{r, warning=FALSE, message=FALSE}
treetib_w_data<-left_join(treetib,metadata,
                          by = c("label"="Assembly.Accession"))
head(treetib_w_data)
```

With these object you can explore and manipulate the phylogeny data, we will go back to this later on.


### treeio full_join

The full_join function of the treeio package allows you to directly bind the metadata to a phylo tree.
```{r, warning=FALSE, message=FALSE}
tree_eco_w_meta<-full_join(tree_eco,metadata,by=c("label"="Assembly.Accession"))

class(tree_eco_w_meta)
tree_eco_w_meta
```

Now you can use this tree object with `ggtree` and reference the table of variants in the styling.

```{r message=FALSE, warning=FALSE}
library(ggtree)
library(RColorBrewer)

plo<-ggtree(tree_eco_w_meta) +
    geom_nodelab(size=2,color="black") +
    geom_tiplab(aes(label=Strain),size=2) +
    geom_tippoint(aes(color=Phylogroup)) 
plo
```
