# Tree object handling {#treeobject}

Trees exist as objects of various forms in R before they are printed to a
visual form. These objects allow us to modify tree data in different ways
to enrich and improve visualization.

## `phylo` objects
Trees are initially imported from Newick files by `read.newick` as `phylo`
objects. These are simple objects which outline tree nodes, edges, and
their labels making them easily accessible.

```{r, warning=FALSE, message=FALSE}
library(treeio)
tree0=read.newick("data/example_phylo_Cmi.nwk")
class(tree0) # simple phylo tree
print(tree0) # simple summary
```

You can find the following information in a tree object
```{r, warning=FALSE, message=FALSE}
names(tree0)
```

`tip.label` has the leaf names.
```{r, warning=FALSE, message=FALSE}
head(tree0$tip.label)
```

`node.label` has the tags in the nodes from the Newick file. In the case of RAxML outputs, these tags are the bootstrap support values.

```{r, warning=FALSE, message=FALSE}
print(head(tree0$node.label))
```

`edge` links leaves and inner nodes to build the tree.
```{r, warning=FALSE, message=FALSE}
print(head(tree0$edge))
```

`edge.length` holds the lengths of the edges according to the phylogenetic
 calculations from RAxML. The edges are counted starting from the number
of leaves + 1.
```{r, warning=FALSE, message=FALSE}
print(head(tree0$edge.length))
```

The advantage of the `phylo` object is that its contents are easy to
modify:
```{r, warning=FALSE, message=FALSE}
tree0a<-tree0
xtip<-which(tree0a$tip.label=="GCF_021952865.1")
tree0a$tip.label[xtip]="changed"
plot(tree0a)
rm(tree0a)
```
## tibbles and tidytree objects
The downside of `phylo` objects is that they are not easy to enrich with
needed annotations. A good way to cover this is `tidytree` objects, which
seamlessly merge all data and make them easily accessible for routines 
downstream. `tidytree` objects, though, are S4 objects which have more
rigid encapsulation.
Building `tidytree` objects from `phylo` objects can become cumbersome
using regular data manipulation routines. Fortunately, tibble functions
make this construction much easier.

```{r, warning=FALSE, message=FALSE}
treetib<-as_tibble(tree0)
class(treetib)
print(treetib)
```

This makes it easy to add metadata. We will use some annotations as example. Remember that metadata labels must be in the same order as tip labels.
```{r message=FALSE, warning=FALSE}
library(dplyr)
# Charge metadata and keep relevant columns
metadata <- readr::read_tsv("data/Experiment_Design_example_1.tsv")
metadata<-metadata %>% select(Assembly.Accession,Strain,
                              Location,Source,Organism)
#Reorder to match phylo tree object
metadata <- metadata %>%
  filter(Assembly.Accession %in% tree0$tip.label) %>%
  arrange(match(Assembly.Accession, tree0$tip.label))

print(head(metadata))
```


## ggtree objects

`ggplot2` produces graphs, and `ggtree` is built on top to specialize in plotting trees. However, to generate the plots `ggplot2` creates objects
which specify plot properties. These objects can be handled in several
ways to create more detailed plots.

We creatte a ggtree plot object
```{r, warning=FALSE, message=FALSE}
library(ggtree)
plo<-ggtree(tree0)
class(plo)
```

And plot it
```{r cmi_phylo_naked, warning=FALSE, message=FALSE}
plo
```


ggplot's added methods operate on the object
```{r, warning=FALSE, message=FALSE}
plo0<-plo +
      geom_nodelab(color="black",breaks=NA) +
      geom_tiplab(aes(label=label),fontface=2)

print(plo$layers) # original object
print(plo0$layers) # new object has added information
```


Another way is to use the `cowplot` library to print the plot as a grid of several objects. This comes useful to create complex plots and legends in
the cases in which `ggplot` makes this difficult.

Lets create a custom coloring scheme for our phylogeny
```{r, warning=FALSE, message=FALSE}
library(cowplot)
library(RColorBrewer)

metadata$Organism<-as.factor(metadata$Organism)
metadata$Source<-as.factor(metadata$Source)

colrs=c(brewer.pal(length(levels(metadata$Organism)),"Paired"),
        "palegreen","chartreuse4")

names(colrs)=c(levels(metadata$Organism),levels(metadata$Source))
colrs
```

Create a data frame for coloring
```{r, warning=FALSE, message=FALSE}
annot<-metadata %>% select(Organism,Source)
rownames(annot)<-metadata$Assembly.Accession
```

POR FAVOR AYÚDENME AQUÍ, NO LOGRO QUE ME SALGA ESTA RECETA
DE FILOGENIA CON HEATMAPS!! El data frame annot cambió con respecto
al código original que creo que es de D

```{r, warning=FALSE, message=FALSE, eval=FALSE}
plo1<-gheatmap(plo0,annot[,1,drop=F]) + 
             scale_color_manual(annot$Organism,values=colrs) + 
             theme(legend.position="right")
plo1


leg1=get_legend(plo1)

plo2=gheatmap(plo0,annot[,"Host",drop=F]) + 
              scale_color_manual("Host",values=colrs) + 
              theme(legend.position="right")
leg2=get_legend(plo2)

plo3=gheatmap(plo0,annot) + 
     scale_color_manual(values=colrs) + 
     theme(legend.position="none")

all_leg=plot_grid(NULL,leg1,leg2,NULL,nrow=1,ncol=4,
                  rel_widths=c(0.25,1,1,0.25))
plot_grid(plo3,all_leg,nrow=2,ncol=1,
          rel_heights=c(1.3,0.15))

```
