
## ggtree objects

`ggplot2` produces graphs, and `ggtree` is built on top to specialize in plotting trees. However, to generate the plots `ggplot2` creates objects
which specify plot properties. These objects can be handled in several
ways to create more detailed plots.

We creatte a ggtree plot object
```{r, warning=FALSE, message=FALSE}
library(ggtree)
plo<-ggtree(tree0)
class(plo)
```

And plot it
```{r cmi_phylo_naked, warning=FALSE, message=FALSE}
plo
```


ggplot's added methods operate on the object

```{r, warning=FALSE, message=FALSE}
plo0<-plo +
      geom_nodelab(color="black",breaks=NA) +
      geom_tiplab(aes(label=label),fontface=2)

print(plo$layers) # original object
print(plo0$layers) # new object has added information
```


Another way is to use the `cowplot` library to print the plot as a grid of several objects. This comes useful to create complex plots and legends in
the cases in which `ggplot` makes this difficult.

Lets create a custom coloring scheme for our phylogeny
```{r, warning=FALSE, message=FALSE}
library(cowplot)
library(RColorBrewer)

metadata$Organism<-as.factor(metadata$Organism)
metadata$Source<-as.factor(metadata$Source)

colrs=c(brewer.pal(length(levels(metadata$Organism)),"Paired"),
        "palegreen","chartreuse4")

names(colrs)=c(levels(metadata$Organism),levels(metadata$Source))
colrs
```

Create a data frame for coloring
```{r, warning=FALSE, message=FALSE}
annot<-metadata %>% select(Organism,Source)
rownames(annot)<-metadata$Assembly.Accession
```

POR FAVOR AYĂDENME AQUĂ, NO LOGRO QUE ME SALGA ESTA RECETA
DE FILOGENIA CON HEATMAPS!! El data frame annot cambiĂł con respecto
al cĂłdigo original que creo que es de D

```{r, warning=FALSE, message=FALSE, eval=FALSE}
plo1<-gheatmap(plo0,annot[,1,drop=F]) + 
             scale_color_manual(annot$Organism,values=colrs) + 
             theme(legend.position="right")
plo1


leg1=get_legend(plo1)

plo2=gheatmap(plo0,annot[,"Host",drop=F]) + 
              scale_color_manual("Host",values=colrs) + 
              theme(legend.position="right")
leg2=get_legend(plo2)

plo3=gheatmap(plo0,annot) + 
     scale_color_manual(values=colrs) + 
     theme(legend.position="none")

all_leg=plot_grid(NULL,leg1,leg2,NULL,nrow=1,ncol=4,
                  rel_widths=c(0.25,1,1,0.25))
plot_grid(plo3,all_leg,nrow=2,ncol=1,
          rel_heights=c(1.3,0.15))

```
